---
title: "Estimation of the cell-type proportions using reference-free method (RefFreeEWAS)"
author:
  
- name: Paulina Jedynak, PhD
  affiliation: Team of Environmental Epidemiology applied to Reproduction and Respiratory Health, Inserm, CNRS, University Grenoble Alpes, Institute for Advanced Biosciences (IAB), U1209 Joint Research Center, Grenoble, France
  email: paulina.jedynak@univ-grenoble-alpes.fr
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: xelatex
    template: svm-latex-ms2.tex
biblio-style: apsr
fontfamily: mathpazo
fontsize: 11pt
geometry: margin = 1in
keywords: DNA methylation, cell-type proportions, ref-free based method, RefFreeEWAS
bibliography: reffree_cell_mix.bib
abstract: This is a reproducible example showing how to apply the reference-free method for estimation of the cell-type proportions in a sample using RefFreeEWAS package [@houseman2016]. A more detailed description of each step and theoretical background is available in the supplementary file reffree_cell_mix_tutorial.pdf to be found at https://github.com/paujedynak/reffree_cell_mix_tutorial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, message = FALSE)
```

****

```{r, warning = FALSE}
# Install and load needed packages
list_packages <- c("RefFreeEWAS", "FactoMineR", "factoextra")
new_packages <- list_packages[!(list_packages %in% installed.packages()[, "Package"])]

if (length(new_packages)) {
  install.packages(new_packages)
}

library(RefFreeEWAS)
library(FactoMineR)
library(factoextra)
```


```{r}
# Load data
data(HNSCC)
print(Y.HNSCC.averageBetas[1:4, 1:4])

# Create exemplary datasets
short <- Y.HNSCC.averageBetas[sample(1:1000, 500), ]
final <- Y.HNSCC.averageBetas[1:1000, ]
```

****

### PART I: Estimate K (the number of cell types in your sample) for a subset of CpG probes

Note: The right choice of K is crucial for a successful estimation of cell proportions. Using a subset of probes enables more precise estimation of K as well as saves computational time. There are different methods of probe selection, e.g. selection based on the probes' variability (e.g. 10,000 most variable probes can be used). Another interesting method is removing the probes whose methylation is affected by factors that are not influencing cell proportions [@decamps2020].   

Below I present an exemplary code to select the probes with both methods:

```{r, eval = FALSE}
# Selecting 10,000 most variable probes
variance <- apply(methylation_matrix, 2, var, na.rm = TRUE)
most_variant_10k <- methylation_matrix[, names(variance[order(variance, decreasing = TRUE)][1:10000])]
```

```{r, eval = FALSE}
# Filtering out the probes whose methylation is associated with factors that are not influencing cell proportions (like: batch effect). For details see Decamps et al. 2020 and package medepir

# Choose factors that are affecting methylation but not influencing cell proportions 
convariates_dataset <- convariates_dataset %>%
  dplyr::select(factor_1, factor_2, factor_3, factor_4)

# Select the probes
selected_probes <- medepir::CF_detection(D = methylation_matrix, exp_grp = convariates_dataset)
```


Selected probes are then used to estimate the K. Again we have (at least) two methods to choose from.    

If we decided to use 10k most variable probes, we can continue with the algorithm proposed by Houseman et al.

```{r}
# Run an estimation for a subset of CpG probes, with different values of K
# Y is the input methylation matrix (in this case it contains 500 CpGs, this would be replaced by 'most_variant_10k').
# Klist represents values of K to be tested (for the purpose of the tutorial I test Ks from 1 to 10)
omega_array <- RefFreeEWAS::RefFreeCellMixArray(Y = short, Klist = 1:10)
```

`RefFreeEWAS` package offers a deviance measure that may help to estimate the right value of K.

```{r}
# Run bootstrap analysis to estimate deviance for each K
# rfArray is a list of RefFreeCellMix objects, Y is the input methylation matrix (in this case it
# contains 500 selected CpGs) and R is a number of bootstrapped vectors to return
set.seed(13446)
testBootDevs <- RefFreeEWAS::RefFreeCellMixArrayDevianceBoots(rfArray = omega_array, Y = short, R = 10)

# To find the best K compare the deviances and choose the minimal one
mean_dev <- apply(testBootDevs[-1, ], 2, mean, trim = 0.25)
which.min(mean_dev)
```

An optimal K is `r which.min(mean_dev)`.

```{r}
# Assign a value of K for further steps
K_optimal_deviance <- which.min(mean_dev)
```


If we decided to use probes remaining after exclusion of those whose methylation is affected by factors that are not influencing cell proportions, we can continue with the algorithm proposed by Decamps et al. They propose a visual inspection of the scree plot (showing the eigenvalues of the selected probes methylation matrix in descending order) as a powerful technique to choose the optimal K. For choosing K, Cattell’s rule is applied, which states that components corresponding to eigenvalues to the left of the straight line should be retained ([@cattell1966]). When the actual number of different cell types is equal to K, we expect that there are (K-1) eigenvalues would correspond to the mixture of cell types and that other eigenvalues would correspond to noise (or other unaccounted for confounders).

```{r}
# Run PCA on selected probes methylation matrix and plot results
# X is the input methylation matrix (in this case it contains 500 CpGs, this would be replaced by 'selected_probes').
pca <- FactoMineR::PCA(X = t(short), graph = FALSE)
factoextra::fviz_screeplot(X = pca, geom = "line")
```

Using Cattlle's rule we choose the optimal K to be 6.

```{r}
# Assign a value of K for further steps
K_optimal_scree <- 6
```

****

### PART II: Estimate cell mix proportion using chosen K

At this step we use the subset of probes (selected by one method or another) and the corresponding optimal K to estimate the final cell mix proportion. There are several algorithms for doing it, in this tutorial I present only the one proposed by Houseman et al. (from the `RefFreeEWAS` package)

Without explicit initialization of the M matrix (recommended). There is an ongoing debate what should be the initialization matrix, as it can significantly influence the results of the cell-type proportion estimation [@decamps2020]. 
One solution would be to obtain cell-type proportions using both methods and compare the deviance of Y matrices reconstructed using each solution with the original input matrix Y, and choosing the cell-type proportion matrix returning lower deviance from original matrix Y.


```{r, eval = FALSE}
# Houseman
# Estimate cell mix proportion without explicit initialization basing on 10k most variable probes
# Y represents the input methylation matrix (this time containing either the subset of probes 'most_variant_10k' or all available CpGs) and K_optimal is the value of K that was chosen in PART I

cell_mix <- RefFreeEWAS::RefFreeCellMix(Y = most_variant_10k, K = K_optimal_deviance, verbose = FALSE)
# OR
cell_mix <- RefFreeEWAS::RefFreeCellMix(Y = methylation_matrix, K = K_optimal_deviance, verbose = FALSE)

```


```{r, eval = FALSE}
# Decamps
# Estimate cell mix proportion without explicit initialization basing on selected probes remaining after removal of probes whose methylation is not associated with cell mix
# Y represents the input methylation matrix (containing the subset of probes 'selected_probes') and K_optimal is the value of K that was chosen in PART I
cell_mix <- RefFreeEWAS::RefFreeCellMix(Y = selected_probes, K = K_optimal_scree, verbose = FALSE)
```


An exemplary estimation of the cell proportions matrix will look as follows

```{r, include = FALSE}
# Estimate cell mix proportion without explicit initialization 
cell_mix <- RefFreeEWAS::RefFreeCellMix(Y = short, K = K_optimal_deviance, verbose = FALSE)

# Print omega (cell mix proportions for each sample). Columns indicate the cell types (from 1 to 6) and rows - individuals
head(cell_mix$Omega)
```


### APPENDIX

Example of use of the `Yfinal` argument.

Some functions from the `RefFreeEWAS` package allow for an optional argument Yfinal. By default, it is set to NULL which makes it equal to Y. We can also set it for a different value. 

In the `RefFreeEWAS` package help we read:    
"Note that the decomposition will be based on Y, but Yfinal (=Y by default) will be used to determine the final value of Mu based on the last iterated value of Omega."

Setting Yfinal different from NULL will cause the following: the matrix decomposition will be based on the information provided by the CpGs from the `Y` argument of the function, therefore Omega and Mu will be estimated basing on `Y`. However, the final displayed solution for Mu will converge to the methylation matrix provided in `Yfinal` argument.

Yfinal = NULL = Y → Y = M(~Y~) x Ω(~Y~)^T^ both M and Ω are estimated basing on Y.    

Yfinal = Y1 → Y1 = M(~Y1~) x Ω(~Y~)^T^ both M and Ω are estimated basing on Y; Returned M will be adjusted to satisfy the new equation Y1 = M(~Y1~) x Ω(~Y~)^T^ is going to converge to.


Let's compare results obtained by the `RefFreeCellMix` without and with `Yfinal` argument:

Yfinal = NULL = Y

```{r}
# Run cell-type proportion estimation without Yfinal argument
# Y is the input methylation matrix, in this case it contains 500 selected CpGs, K_optimal is the value of K that was chosen in PART II and Yfinal is NULL so by defaul it equals Y
cell_mix_Y <- RefFreeEWAS::RefFreeCellMix(Y = short, K = K_optimal_deviance, Yfinal = NULL, verbose = FALSE)

# Print omega and mu for estimation without Yfinal argument
head(cell_mix_Y$Omega)
dim(cell_mix_Y$Omega)
head(cell_mix_Y$Mu)
dim(cell_mix_Y$Mu)
```


Now let's add `Yfinal` argument:

Yfinal = final

```{r}
# Run cell-type proportion estimation with Yfinal argument
# Y is the input methylation matrix, in this case it contains 500 selected CpGs, K_optimal is the value of K that was chosen in PART II and Yfinal is the methylation matrix that the final solution of Mu will converge to
cell_mix_Yfinal <- RefFreeEWAS::RefFreeCellMix(Y = short, K = K_optimal_deviance, Yfinal = final, verbose = FALSE)

# Print omega and mu for estimation with Yfinal argument
head(cell_mix_Yfinal$Omega)
dim(cell_mix_Yfinal$Omega)
head(cell_mix_Yfinal$Mu)
dim(cell_mix_Yfinal$Mu)
```


For the two solutions, Omega stays the same and Mu is changing after adding `Yfinal` argument, in order to conform to the Yfinal = final.

```{r}
sessionInfo()
```

### References:

